---
title: "Phytoalexins"
author: "Hauschild et al. 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)

library(tidyverse)
library(lme4)
library(emmeans)
library(lmerTest)
library(ggpubr)
```


# Data import

```{r eval=FALSE}
# Load packages
library(tidyverse)
library(lme4)
library(emmeans)
library(lmerTest)
library(ggpubr)
```

```{r}
dat <- read.csv2("WP3_Phytoalexine_t8.csv",
                 header=TRUE,
                 strip.white = TRUE,
                 sep=";",
                 dec=".")
str(dat)

dat$Table <- "1"
dat$Table[which(dat$Block=="C")] <- "2"
dat$Table[which(dat$Block=="D")] <- "2"
```

# Grafical overview

```{r}
ggplot(dat, aes(x=Treatment,
                y=Phytoalexin))+
        theme_bw()+
        geom_jitter(aes(color=Genotype),
                    width=0.2,
                    alpha=0.7)+
        facet_grid(Table+Block~Soil)
```

# Modelling


```{r}
# Model on response scale
fit_resp <- lmer(Phytoalexin ~ Treatment*Genotype* Soil+
                    (1|Table) + (1|Table:Block)+
                    (1|Table:Treatment) + (1|Table:Block:Treatment)+
                    (1|Table:Genotype) + (1|Table:Block:Genotype)+
                    (1|Table:Soil) + (1|Table:Block:Soil),
            data=dat)
```

Residual variance increases with fitted values.
```{r}
plot(fit_resp)
```


```{r}
# Add 1 to each observation to overcome log(0)
dat$Phyto1 <- dat$Phytoalexin+1

# Model on log1p-scale
fit <- lmer(log(Phyto1) ~ Treatment*Genotype* Soil+
                    (1|Table) + (1|Table:Block)+
                    (1|Table:Treatment) + (1|Table:Block:Treatment)+
                    (1|Table:Genotype) + (1|Table:Block:Genotype)+
                    (1|Table:Soil) + (1|Table:Block:Soil),
            data=dat)
```

The zeros (ones) are clearly visible in the residuals vs. fitted plot, but residual
variance seems to be relatively homogeneous.
```{r}
plot(fit)
```

Several interactions between main effects and randomization structures play a role.
```{r}
print(VarCorr(fit) ,comp=c("Variance","Std.Dev."), digits=2)
```

# ANOVA

```{r}
anova(fit, ddf = "Kenward-Roger")
```

The treatment:soil interaction is significant as well as the interaction between
genotype and soil.

# Mean comparisons

## Treatment soil interaction

```{r cache=TRUE}
# simple contrasts treatment + soil:treatment effects
comp_ts_simple <- emmeans(fit,
                   specs="Treatment",
                   by="Soil",
                   contr="pairwise",
                   adjust="mvt",
                   type="response")
```

```{r}
# Means on response scale
means_ts_simple <- data.frame(comp_ts_simple$emmeans) %>% 
        mutate(response=response-1)

means_ts_simple
```

```{r cache=TRUE}
# Conrtasts on y+1 scale
contr_ts_simple <- data.frame(confint(comp_ts_simple$contrasts))
contr_ts_simple
```


```{r cache=TRUE}
# interaction contrasts
means_ts_inter <- emmeans(fit, 
                         specs=c("Treatment", "Soil"))

contr_ts_inter <- contrast(means_ts_inter,
                       interaction=c(Treatment="pairwise",
                                     Soil="pairwise"),
                       adj="mvt",
                       type="response") %>% 
        confint() %>%
        data.frame() %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               lower.CL=round(lower.CL, 2),
               upper.CL=round(upper.CL, 2))

contr_ts_inter
```


```{r echo=FALSE, fig.width=10, fig.height=10, fig.cap = "\\label{fig_ts_inter} Phytoalexines for the genotype soil interaction. A: Average phytoalexin concentration for each genotypek. B: Simple contrasts. C: Interaction contratsts.  Black bars: 95 % confidence intervals (mvt-adjusted for the mean differences in B and C). Mean comparisons are calculated as mean differences on log(y+1) scale. Mean proprtions in B and C are given on y+1 scale. Horizontal line: Visualisation of the null hypothesis."}
means_ts_plot <- ggplot(dat, aes(x=Treatment,
                y=Phytoalexin))+
        theme_bw()+
        geom_jitter(aes(color=Genotype),
                    width=0.2,
                    alpha=0.7)+
        geom_pointrange(data=means_ts_simple,
                        aes(x=Treatment,
                            y=response,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        facet_grid(~Soil)+
        ggtitle("A: Average phytoalexin content within treatments (on response scale)")

plot_ts_simple <- ggplot(contr_ts_simple,
                         aes(x=contrast,
                             y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept = 1)+
        facet_grid(Soil~.)+
        scale_y_continuous(trans="log10")+
        coord_flip()+
        ggtitle("B: Simple contrasts between treatments \n within genotypes (on y+1 scale)")
        
        

plot_ts_inter <- ggplot(contr_ts_inter,
                        aes(x=Soil_pairwise,
                            y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept = 1)+
        facet_grid(Treatment_pairwise~.)+
        scale_y_continuous(trans="log10")+
        coord_flip()+
        ggtitle("C: Interaction contrasts (on y+1 scale)")

ggarrange(means_ts_plot,
          ggarrange(plot_ts_simple,
                    plot_ts_inter,
                    widths = c(2:2.8)),
          nrow=2)
```

\newpage

## Genotype soil interaction

```{r cache=TRUE}
# simple contrasts genotype + soil:genotype effects
comp_gs_simple <- emmeans(fit,
                   specs="Genotype",
                   by="Soil",
                   contr="pairwise",
                   adjust="mvt",
                   type="response")
```

```{r}
# Means on response scale
means_gs_simple <- data.frame(comp_gs_simple$emmeans) %>% 
        mutate(response=response-1)

means_gs_simple
```

```{r cache=TRUE}
# Conrasts on y+1 scale
contr_gs_simple <- data.frame(confint(comp_gs_simple$contrasts))
contr_gs_simple
```


```{r cache=TRUE}
# interaction contrasts
means_gs_inter <- emmeans(fit, 
                         specs=c("Genotype", "Soil"))

contr_gs_inter <- contrast(means_gs_inter,
                       interaction=c(Genotype="pairwise",
                                     Soil="pairwise"),
                       adj="mvt",
                       type="response") %>% 
        confint() %>%
        data.frame() %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               lower.CL=round(lower.CL, 2),
               upper.CL=round(upper.CL, 2))

contr_gs_inter
```


```{r echo=FALSE, fig.width=10, fig.height=10, fig.cap = "\\label{fig_gs_inter} Phytoalexines for the genotype soil interaction. A: Average phytoalexin concentration for each genotypek. B: Simple contrasts. C: Interaction contratsts.  Black bars: 95 % confidence intervals (mvt-adjusted for the mean differences in B and C). Mean comparisons are calculated as mean differences on log(y+1) scale. Mean proprtions in B and C are given on y+1 scale. Horizontal line: Visualisation of the null hypothesis."}
means_gs_plot <- ggplot(dat, aes(x=Genotype,
                y=Phytoalexin))+
        theme_bw()+
        geom_jitter(aes(color=Treatment),
                    width=0.2,
                    alpha=0.7)+
        geom_pointrange(data=means_gs_simple,
                        aes(x=Genotype,
                            y=response,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        facet_grid(~Soil)+
        ggtitle("A: Average phytoalexin content within soils (on response scale)")

plot_gs_simple <- ggplot(contr_gs_simple,
                         aes(x=contrast,
                             y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept = 1)+
        facet_grid(Soil~.)+
        scale_y_continuous(trans="log10")+
        coord_flip()+
        ggtitle("B: Simple contrasts between Genotypes \n within soils (on y+1 scale)")
        
        

plot_gs_inter <- ggplot(contr_gs_inter,
                        aes(x=Soil_pairwise,
                            y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept = 1)+
        facet_grid(Genotype_pairwise~.)+
        scale_y_continuous(trans="log10")+
        coord_flip()+
        ggtitle("C: Interaction contrasts (on y+1 scale)")

ggarrange(means_gs_plot,
          ggarrange(plot_gs_simple,
                    plot_gs_inter,
                    widths = c(2:2.8)),
          nrow=2)
```

