---
title: "Shannon index for fungi"
author: "Hauschild et al. 2024"
date: "2024-12-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message=FALSE)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(emmeans)
library(nlme)
library(ggpubr)
```

# Data import

```{r eval=FALSE}
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(emmeans)
library(nlme)
library(ggpubr)
```


```{r}
# Data import
dat <- read.csv2("df_psO_WP3_ITS_alpha_estimates.csv",
                 sep=";",
                 dec=".") %>% 
        mutate(replicate=factor(replicate),
               table="1",
               Evenness=Shannon/log(Observed))

# blocks c and d were placed on table 2
dat$table[which(dat$block=="C")] <- "2"
dat$table[which(dat$block=="D")] <- "2"

str(dat)
```


# Grafical overview

```{r cache=TRUE, echo=FALSE}
# Grafical overview

ggplot(dat, aes(x=rootstock, y=Shannon))+
        theme_bw()+
        geom_jitter(aes(fill = Evenness,
                        shape=soil_type),
                    width=0.2,
                    size=3,
                    alpha=0.9)+
        facet_nested(table+block~site)+
        scale_shape_manual(values=c(21,22,23))+
        scale_fill_gradient2(low = "blue", 
                            mid = "white",
                            high = "red",
                            midpoint = 0.725,
                            limits=c(0.50, 0.95))


```

# Modelling

## Linear mixed model with nlme::lme

In order to be in line with the analysis regarding the Shannon indices for the bacterial
community, also the alpha diversity of the fungi community was modelled based
on `nlme::lme()`. The same mixed model was fit, to allow for heterogeneous residual
variance between treatment groups. Since `lme()` only allows a hierarchical structure
in the random effects, the tables, bocks nested in tables and the interaction 
rootstock:site nested in blocks and tables was models as random factors.

```{r cache=TRUE}
# Combination of rootstock and site
dat$rs_inter <- droplevels(factor(dat$rootstock) : factor(dat$site))

# Fit the model
fit_nlme <- lme(fixed = Shannon ~ rootstock + site + rootstock:site,
                data = dat, 
                random = ~ 1|table/block/rs_inter,
                weights = varIdent(form = ~ 1|rootstock*site),
                method = "REML")
```

All random effects play a role, but residual variance is dominating.
```{r}
VarCorr(fit_nlme)
```

Residual variance is homogeneous.
```{r}
plot(fit_nlme)
```

# ANOVA

```{r}
anova(fit_nlme)
```

The interaction between rootstock and site is significant.


# Multiple mean comparisons for the two way interaction

```{r cache=TRUE}
# Simple contrasts
comp_rs <- emmeans(fit_nlme,
                   specs = "rootstock",
                   by="site",
                   contr = "pairwise",
                   adjust = "mvt",
                   type="response")
```

```{r}
means_rs <- data.frame(comp_rs$emmeans)
means_rs
```


```{r}
ggplot(dat, aes(x=rootstock, y=Shannon))+
        theme_bw()+
        geom_jitter(aes(fill = Evenness,
                        shape=soil_type),
                    width=0.2,
                    size=3,
                    alpha=0.9)+
        geom_pointrange(data=means_rs,
                        aes(y=emmean,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        facet_nested(~site)+
        scale_shape_manual(values=c(21,22,23))+
        scale_fill_gradient2(low = "blue", 
                            mid = "white",
                            high = "red",
                            midpoint = 0.725,
                            limits=c(0.50, 0.95))
```

```{r}
contr_rs <- data.frame(comp_rs$contrasts)
contr_rs
```

```{r}
plot(confint(comp_rs$contrasts))+
        theme_bw()+
        geom_vline(xintercept=1)
```

```{r}
# Interaction contrasts
means_inter_rs <- emmeans(fit_nlme,
                       specs= ~ rootstock:site,
                       type="response")

means_inter_rs

```

```{r}
# Interaction contrasts
comp_inter_rs <- contrast(means_inter_rs,
                       interaction=c(rootstock="pairwise", 
                                     site="pairwise"),
                       adj="mvt",
                       type="response") 

comp_inter_rs
```


```{r}
plot(confint(comp_inter_rs))+
        theme_bw()+
        geom_vline(xintercept=1)
```










