---
title: "Root fresh mass"
author: "Hauschild et al. 2024"
date: "2024-11-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
library(lme4)
library(ggpubr)
library(lmerTest)
library(emmeans)
```

# Data import

```{r eval=FALSE}
# Load necessary packages 
library(tidyverse)
library(lme4)
library(ggpubr)
library(lmerTest)
library(emmeans)
```

The data set WP3.csv is available at the **Bonares repo xy (insert link or DOI)**
```{r}
# Data import
dat <- read.csv2("WP3.csv",
                 strip.white = TRUE) %>% 
        mutate(Genotype=factor(Genotype),
               Soil=factor(Soil),
               Treat=factor(Treat, levels=c("ARD", "G")),
               Plant=factor(Plant),
               Block=factor(Block),
               Table=factor(Table)) %>% 
        na.omit()

# Overview about data set
str(dat)
```
\newpage

# Linear mixed model

A linear mixed model is fit in which the ln-transformed root fresh mass depends on 
the soil, the different genotypes, the different treatments and their possible 
two-way and three-way interactions as fixed effects. The tables, the blocks and
their interactions with the main effects were modeled as random effects.
```{r}
# Fit the model
fit_RFM <- lmer(log(RFM) ~ Soil*Genotype*Treat + 
                        (1|Table) + 
                        (1|Table:Block)+
                        (1|Table:Soil)+ (1|Table:Genotype)+ (1|Table:Treat)+
                        (1|Table:Block:Soil)+ (1|Table:Block:Genotype)+
                        (1|Table:Block:Treat), 
                data=dat)
```

### Model diagnostic
```{r}
# Residuals vs. fitted plot
plot(fit_RFM)
```

The residuals are evenly distributed around the 0. It seems that assumptions are med.
\newpage

### Variance components

```{r}
print(VarCorr(fit_RFM) ,comp=c("Variance","Std.Dev."), digits=2)
```

Residual variance explains roughly 97 % of the total variance in the data, meaning
that there were only few deviations between the blocks (which explain roughly 
0.008 % of the total variance.)


## ANOVA

Type 3 Anova with K-R approximation of degrees of freedom.

```{r cache=TRUE}
anova(fit_RFM, ddf = "Kenward-Roger")
```

The treatment (ARD substrate vs.  gamma irradiated substrate) alone explains roughly
two thirds of the sum of squares and therefore having the strongest effect on 
root fresh mass. Compared to the treatment, all other model terms (including the 
interactions) are of minor importance, but are also significant (except the three
way interaction).
This means that at least two genotypes differ with respect to their root fresh mass 
depending on the soil, at least two genotypes react differently with respect to 
the treatment and the fresh mass per treatment depends on the soils as well, but
no differences between the treatment can be found in each combination of the 
six soils and the six genotypes.


\newpage 

# Multiple mean comparisons

All multiple mean comparisons were calculated on ln-scale, but back-transformed to
response scale for easier interpretation. Back transformation changes the hypothesis
$H_0: ln(\mu_i) - ln(\mu_{i'}) = 0$ to $H_0: exp(ln(\mu_i) - ln(\mu_{i'})) = \mu_i / \mu_{i'} = 1$.
All contrast tests were adjusted using the multivariate t-distribution following
Hothorn et al. 2008. 

## Treatment genotype interaction

```{r cache=TRUE}
# Treatments within genotypes
# Simple contrasts: Main effect for treatment (ARD/G)  + treatment:genotype effects 
# averaged over soils

comp_RFM_treat <- emmeans(fit_RFM, 
                    specs="Treat", 
                    by="Genotype",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```

```{r}
# Means and their pointwise 95% confidence intervals
means_RFM_treat <- data.frame(comp_RFM_treat$emmeans)
means_RFM_treat 
```


```{r}
# Contrasts, p-values etc.
contr_RFM_treat <- data.frame(comp_RFM_treat$contrasts)
contr_RFM_treat
```
\newpage


```{r cache=TRUE}
# Interaction contrasts (ARD / G) between genotypes
# estimates, p-values etc.
inter_means_treat <- emmeans(fit_RFM, specs=c("Treat", "Genotype"))

inter_cont_treat <- contrast(inter_means_treat,
                       interaction=c(Treat="pairwise", Genotype="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_treat
```

\newpage

```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (ARD / G) between genotypes
inter_cont_ci_treat <- confint(inter_cont_treat) %>% 
        data.frame() %>% 
        mutate(treat_pw=Treat_pairwise,
               genotype_pw=Genotype_pairwise,
               ratio=round(ratio, 3),
               lower.CL=round(lower.CL, 3),
               upper.CL=round(upper.CL, 3)) %>% 
        select(treat_pw,    
               genotype_pw,
               ratio,
               lower.CL,
               upper.CL)

inter_cont_ci_treat
```

### Overview about results

The analysis of the genotype-treatment interaction is summarized in Fig. \ref{fig_geno_treat}.
Fig \ref{fig_geno_treat}A provides and overview about the raw data, the average 
root fresh mass per treatment within each genotype and corresponding pointwise 95 % confidence
intervals.
The proportional change of root freshmass between ARD and gamma irradiated substrate
for each genotyope is given in fig \ref{fig_geno_treat}A together with
multiplicity adjusted 95 % confidence intervals. Since the null hypothesis
($H_0: \mu_i / \mu_{i'} = 1$) is covered by none of the confidence intervals,
it can be deduced that in each genotype, the root fresh mass for ARD infected 
soil is significantly lower than in gamma irradiated soil. However, the magnitude
of root fresh mass reduction (ARD/G) differs only between genotypes M26 and AL0739
since its adjusted 95 % confidence interval does not include the $H_0: \mu_i / \mu_{i'} = 1$
(\ref{fig_geno_treat}C).

```{r echo=FALSE, cache=TRUE, fig.width=10, fig.height=9, fig.cap = "\\label{fig_geno_treat} Analysis of the genotype treatment interaction. A: Treatment means within genotypes. B: Proportion of root fresh mass between the treatments (ARD / gARD) for each genotype. C: Comparison of fresh mass proportions between genotypes. Black bars: 95 % confidence intervals (mvt-adjusted for the tests visualised in B and C)."}
# Graphical display of the results

# Observations, means, 95% pointwise CI
plot_RFM_1_treat <- ggplot(dat, aes(x=Treat, y=RFM))+
        theme_bw()+
        geom_jitter(aes(color=Soil),
                    width=0.2,
                    alpha=0.3)+
        geom_pointrange(data=means_RFM_treat,
                        aes(y=response,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        facet_grid( ~ Genotype)+
        # scale_color_manual(values=c("Hg"="#DFA398",
        #                             "Ho"="#883b41",
        #                             "El"="#41917C",
        #                             "Pi"="#A9D284",
        #                             "Ru"="#505778",
        #                             "Me"="#A0A5C0"))+
        ylab("Root fresh mass [g]")+
        xlab("Treatment")+
        ggtitle("A: Treatment means")


# Simultaneous 95 % CI for contrast G- ARD within each genotype
ci_SMF_treat <- data.frame(confint(comp_RFM_treat$contrasts))

plot_RFM_2_treat <- ggplot(ci_SMF_treat, aes(x=Genotype))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept=1)+
        xlab("Contrast")+
        ylab("Estimate")+
        coord_flip()+
        ggtitle("B: ARD/G within genotypes")


plot_RFM_3_treat <- ggplot(inter_cont_ci_treat, aes(x=genotype_pw, y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        geom_hline(yintercept=1)+
        ylab("Estimate")+
        xlab("Contrast")+
        coord_flip()+
        ggtitle("C: ARD/G between genotypes")


ggarrange(plot_RFM_1_treat,
          ggarrange(plot_RFM_2_treat,
                    plot_RFM_3_treat,
                    widths = c(2:2.8)),
          nrow=2)
```

\newpage

## Soil treatment interaction

```{r cache=TRUE}
# Treatments within soils
# Simple contrasts: Main effect for treatment (ARD/G)  + treatment:soil effects 
# averaged over genotypes

comp_RFM_soil <- emmeans(fit_RFM, 
                    specs="Treat", 
                    by="Soil",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```

```{r}
# Means and their pointwise 95% confidence intervals
means_RFM_soil <- data.frame(comp_RFM_soil$emmeans)
means_RFM_soil
```


```{r}
# Contrasts, p-values etc.
contr_RFM_soil <- data.frame(comp_RFM_soil$contrasts)
contr_RFM_soil
```

\newpage

```{r cache=TRUE}
# Interaction contrasts (ARD / G) between genotypes
# estimates, p-values etc.
inter_means_soil <- emmeans(fit_RFM, specs=c("Treat", "Soil"))

inter_cont_soil <- contrast(inter_means_soil,
                       interaction=c(Treat="pairwise", Soil="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_soil
```

\newpage

```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (ARD / G) between genotypes
inter_cont_ci_soil <- confint(inter_cont_soil) %>% 
        data.frame() %>% 
        mutate(treat_pw=Treat_pairwise,
               soil_pw=Soil_pairwise,
               ratio=round(ratio, 3),
               lower.CL=round(lower.CL, 3),
               upper.CL=round(upper.CL, 3)) %>% 
        select(treat_pw,    
               soil_pw,
               ratio,
               lower.CL,
               upper.CL)

inter_cont_ci_soil
```

### Overview about results

The analysis of the soil-treatment interaction is summarized in Fig. \ref{fig_soil_treat}.
Fig \ref{fig_soil_treat}A provides an overview about the raw data, the average 
root fresh mass per treatment within each soil and corresponding pointwise 95 % confidence
intervals.
The proportional change of root freshmass between ARD and gamma irradiated substrate
for each soil is given in fig \ref{fig_soil_treat}B together with
multiplicity adjusted 95 % confidence intervals. Since the null hypothesis
($H_0: \mu_i / \mu_{i'} = 1$) is covered by none of the confidence intervals,
it can be deduced that in each soil, the root fresh mass for ARD infected 
substrate is significantly lower than in its gamma irradiated counterpart.
The magnitude of root fresh mass reduction (ARD/G) is significantly bigger in 
Pinneberg than in Ruthe and significantly bigger in \textbf{Co?} than in Heidgraben,
because the corresponding confidence intervals are relatively far above the 1 
(\ref{fig_soil_treat}C). It is unclear if the magnitude of root fresh mass reduction
differs between Me and Ruthe as well as between Co and Ruthe and between CO and EL,
because the Lower border of the corresponding confidence intervals is pretty close
to the null hypothesis of 1. 
The magnitude of root fresh mass reduction is significantly smaller in HE than in Pi, 
in He than in Me, in El than in PI, because the complete confidence interval is lower than 
the null hypothesis of 1. Also the magnitude of root fresh mass reduction in El
appears to be significantly smaller than in Me, but the corresponding upper confidence 
limit is relatively close to 1. 

```{r echo=FALSE, cache=TRUE, fig.width=10, fig.height=9, fig.cap = "\\label{fig_soil_treat} Analysis of the soil treatment interaction. A: Treatment means within soils. B: Proportional change of root fresh mass between the treatments (ARD / gARD) for each soil. C: Comparison of fresh mass proportions between soils. Black bars: 95 % confidence intervals (mvt-adjusted for the tests visualised in B and C)."}
# Graphical display of the results

# Observations, means, 95% pointwise CI
plot_RFM_1_soil <- ggplot(dat, aes(x=Treat, y=RFM))+
        theme_bw()+
        geom_jitter(aes(color=Genotype),
                    width=0.2,
                    alpha=0.3)+
        geom_pointrange(data=means_RFM_soil,
                        aes(y=response,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        facet_grid( ~ Soil)+
        # scale_color_manual(values=c("Hg"="#DFA398",
        #                             "Ho"="#883b41",
        #                             "El"="#41917C",
        #                             "Pi"="#A9D284",
        #                             "Ru"="#505778",
        #                             "Me"="#A0A5C0"))+
        ylab("Root fresh mass [g]")+
        xlab("Treatment")+
        ggtitle("A: Treatment means")


# Simultaneous 95 % CI for contrast G- ARD within each genotype
ci_SMF_soil <- data.frame(confint(comp_RFM_soil$contrasts))

plot_RFM_2_soil <- ggplot(ci_SMF_soil, aes(x=Soil))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept=1)+
        xlab("Contrast")+
        ylab("Estimate")+
        coord_flip()+
        ggtitle("B: ARD/G within genotypes")


plot_RFM_3_soil <- ggplot(inter_cont_ci_soil, aes(x=soil_pw, y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        geom_hline(yintercept=1)+
        ylab("Estimate")+
        xlab("Contrast")+
        coord_flip()+
        ggtitle("C: ARD/G between genotypes")


ggarrange(plot_RFM_1_soil,
          ggarrange(plot_RFM_2_soil,
                    plot_RFM_3_soil,
                    widths = c(2:2.8)),
          nrow=2)
```

\newpage

## Soil Genotype interaction


```{r cache=TRUE}
# Treatments within soils
comp_RFM_sg <- emmeans(fit_RFM, 
                    specs="Genotype", 
                    by="Soil",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```

```{r}
# Means and their pointwise 95% confidence intervals
means_RFM_sg <- data.frame(comp_RFM_sg$emmeans)
means_RFM_sg
```


```{r}
# Contrasts, p-values etc.
contr_RFM_sg <- data.frame(comp_RFM_sg$contrasts)
contr_RFM_sg %>% 
        mutate(ratio=round(ratio, 3),
               SE=round(SE, 3),
               df=round(df, 3),
               t.ratio=round(t.ratio, 3),
               p.value=round(p.value, 3))
```

\newpage

```{r cache=TRUE}
# Interaction contrasts (soil_i / soil_i') between genotypes
# estimates, p-values etc.
inter_means_sg <- emmeans(fit_RFM, specs=c("Genotype", "Soil"))

inter_cont_sg <- contrast(inter_means_sg,
                       interaction=c(Genotype="pairwise", Soil="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_sg
```



\newpage

```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (soil_i / soil_i') between genotypes
# Takes very long computing time
# inter_cont_ci_sg <- confint(inter_cont_sg) %>%
#         data.frame() %>%
#         mutate(treat_pw=Treat_pairwise,
#                soil_pw=Soil_pairwise,
#                ratio=round(ratio, 3),
#                lower.CL=round(lower.CL, 3),
#                upper.CL=round(upper.CL, 3)) %>%
#         select(treat_pw,
#                soil_pw,
#                ratio,
#                lower.CL,
#                upper.CL)
# 
# inter_cont_ci_soil
```

# Threeway interaction (Soil:Genotype:Treat)

```{r cache=TRUE}
# Simple effects: Main effect for treatment (ARD/G) + treatment:soil effects + 
# treatment:genotype effects

comp_RFM_3 <- emmeans(fit_RFM, 
                    specs="Treat", 
                    by=c("Soil", "Genotype"),
                    contr="pairwise",
                    adjust="mvt",
                    type="response")

contr_3 <- data.frame(comp_RFM_3$contrasts)

contr_3 %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               p.value=round(p.value, 4)) %>% 
        select(contrast,
               Soil,
               Genotype,
               ratio,
               SE,
               df,
               p.value)
```

```{r fig.width=10, fig.height=10, fig.cap = "\\label{fig_3_way} Estimated proportions for ARD/G for each combination of soil and genotype."}
contr_3_ci <- data.frame(confint(comp_RFM_3$contrasts))

ggplot(contr_3_ci,aes(x=contrast))+
  theme_bw()+
  geom_pointrange(aes(y=ratio,
                      ymin=lower.CL,
                      ymax=upper.CL))+
  facet_grid(Genotype~Soil)+
  geom_hline(yintercept=1)+
  xlab("Contrast")+
  ylab("Estimate")
```

\newpage

In all combinations of the soils and genotypes plants grown in ARD infected substrate
yield less root fresh mass than plants in gamma irradiated substrate (fig. \ref{fig_3_way}). All differences
between ARD and gamma irradiated substrate are significant, since all confidence intervals
do not cover the 1. Also none of the interaction contrasts given below is significant.
Hence, the reduction of root fresh mass caused by the ARD infection (compared to gamma 
irradiated substrate) is constant for all genotype-soil combinations and the
three-way interaction is found to be absent (as already shown in the ANOVA table).


```{r cache=TRUE}
# Interaction contrasts
means_3 <- emmeans(fit_RFM,
                   specs= ~ Soil:Genotype:Treat,
                    type="response")

inter_cont_3 <- contrast(means_3,
                       interaction=c(Genotype="pairwise", 
                                     Soil="pairwise",
                                     Treat="pairwise"),
                       adj="mvt",
                       type="response") 

inter_cont_3_df <- data.frame(inter_cont_3)

inter_cont_3_df %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               p.value=round(p.value, 2)) %>% 
        select(Treat_pairwise,
               Soil_pairwise,
               Genotype_pairwise,
               ratio,
               SE,
               df,
               p.value)

```


