---
title: "Shoot fresh mass"
author: "Hauschild et al. 2024"
date: "2024-11-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
library(lme4)
library(ggpubr)
library(lmerTest)
library(emmeans)
```

# Data import

```{r eval=FALSE}
# Load necessary packages 
library(tidyverse)
library(lme4)
library(ggpubr)
library(lmerTest)
library(emmeans)
```

The data set WP3.csv is available at the **Bonares repo xy (insert link or DOI)**
```{r}
# Data import
dat <- read.csv2("WP3.csv",
                 strip.white = TRUE) %>% 
        mutate(Genotype=factor(Genotype),
               Soil=factor(Soil),
               Treat=factor(Treat, levels=c("ARD", "G")),
               Plant=factor(Plant),
               Block=factor(Block),
               Table=factor(Table)) %>% 
        na.omit()

# Overview about data set
str(dat)
```
\newpage

# Linear mixed model

A linear mixed model is fit in which the ln-transformed shoot fresh mass depends on 
the soil, the different genotypes, the different treatments and their possible 
two-way and three-way interactions as fixed effects. The tables, the blocks and
their interactions with the main effects were modeled as random effects.
```{r}
# Fit the model on log scale
fit_SFM <- lmer(log(SFM) ~ Soil*Genotype*Treat + 
                        (1|Table) + 
                        (1|Table:Block)+
                        (1|Table:Soil)+ (1|Table:Genotype)+ (1|Table:Treat)+
                        (1|Table:Block:Soil)+ (1|Table:Block:Genotype)+
                        (1|Table:Block:Treat), 
                data=dat)

# Fit model on response scale
fit_SFM_resp <- lmer(SFM ~ Soil*Genotype*Treat + 
                        (1|Table) + 
                        (1|Table:Block)+
                        (1|Table:Soil)+ (1|Table:Genotype)+ (1|Table:Treat)+
                        (1|Table:Block:Soil)+ (1|Table:Block:Genotype)+
                        (1|Table:Block:Treat), 
                data=dat)
```

### Model diagnostic
```{r}
# Residuals vs. fitted plot for model on log-scale
plot(fit_SFM)
```

```{r}
# Residuals vs. fitted plot for model on log-scale
plot(fit_SFM_resp)
```

Residuals in the model on response scale are slightly variance heterogene (residuals
for small fitted values tend to be relatively small). Ln-transformation yields
to relatively small residuals for higher fitted values.


### Variance components

```{r}
print(VarCorr(fit_SFM), comp=c("Variance","Std.Dev."), digits=2)
```

Residual variance explains roughly 99.9 % of the total variance in the data, meaning
that all other variance components plan no or only a minir role and the observations
are almost uncorrelated.


## ANOVA

Type 3 Anova with K-R approximation of degrees of freedom.

```{r cache=TRUE}
anova(fit_SFM, ddf = "Kenward-Roger")
```

The treatment (ARD substrate vs.  gamma irradiated substrate) alone explains roughly
43 % of the sum of squares and therefore having the strongest effect on 
shoot fresh mass. The genotypes explain roughly 34 % of the total sum of squares.
Hence this two factors play the major role in the data and all other model terms 
(including the interactions) are of minor importance, but are also significant 
(except for the three way interaction).
This means that at least two genotypes differ with respect to their shoot fresh mass 
depending on the soil, at least two genotypes react differently with respect to 
the treatment and the fresh mass per treatment depends on the soils as well, but
no differences between the treatment can be found in each combination of the 
six soils and the six genotypes.

# Multiple mean comparisons

All multiple mean comparisons were calculated on ln-scale, but back-transformed to
response scale for easier interpretation. Back transformation changes the hypothesis
$H_0: ln(\mu_i) - ln(\mu_{i'}) = 0$ to $H_0: exp(ln(\mu_i) - ln(\mu_{i'})) = \mu_i / \mu_{i'} = 1$.
All contrast tests were adjusted using the multivariate t-distribution following
Hothorn et al. 2008. 

## Treatment genotype interaction

```{r cache=TRUE}
# Treatments within genotypes
comp_SFM_treat <- emmeans(fit_SFM, 
                    specs="Treat", 
                    by="Genotype",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```


```{r}
# Means and their pointwise 95% confidence intervals
means_SFM_treat <- data.frame(comp_SFM_treat$emmeans)
means_SFM_treat 
```


```{r}
# Contrasts, p-values etc.
contr_SFM_treat <- data.frame(comp_SFM_treat$contrasts)
contr_SFM_treat
```


```{r cache=TRUE}
# Interaction contrasts (ARD / G) between genotypes
# estimates, p-values etc.

# AAAAAACCCCCCHHHHHTTTTUUUUUNNNNNGGGGG!!!!!
# Auf richtige Interpretation achten!!
# Auch bei rfm
inter_means_treat <- emmeans(fit_SFM, specs=c("Treat", "Genotype"))

inter_cont_treat <- contrast(inter_means_treat,
                       interaction=c(Treat="pairwise", Genotype="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_treat
```


```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (ARD / G) between genotypes
inter_cont_ci_treat <- confint(inter_cont_treat) %>% 
        data.frame() %>% 
        mutate(treat_pw=Treat_pairwise,
               genotype_pw=Genotype_pairwise,
               ratio=round(ratio, 3),
               lower.CL=round(lower.CL, 3),
               upper.CL=round(upper.CL, 3)) %>% 
        select(treat_pw,    
               genotype_pw,
               ratio,
               lower.CL,
               upper.CL)

inter_cont_ci_treat
```

### Overview about results

The analysis of the genotype-treatment interaction is summarized in Fig. \ref{fig_geno_treat}.
Fig \ref{fig_geno_treat}A provides and overview about the raw data, the average 
shoot fresh mass per treatment within each genotype and corresponding pointwise 95 % confidence
intervals.
The proportional change of shoot freshmass between ARD and gamma irradiated substrate
for each genotyope is given in fig \ref{fig_geno_treat}B together with
multiplicity adjusted 95 % confidence intervals. Since the null hypothesis
($H_0: \mu_i / \mu_{i'} = 1$) is covered by none of the confidence intervals,
it can be deduced that in each genotype, the shoot fresh mass for ARD infected 
soil is significantly lower than in gamma irradiated soil. However, the magnitude
of shoot fresh mass reduction (ARD/G) differs only between genotypes MAL0130 and MAL0739
as well as AR10-3-9 and MAL0739, since its adjusted 95 % confidence interval does
not include the $H_0: \mu_i / \mu_{i'} = 1$ (\ref{fig_geno_treat}C).

```{r echo=FALSE, cache=TRUE, fig.width=10, fig.height=9, fig.cap = "\\label{fig_geno_treat} Analysis of the genotype treatment interaction. A: Treatment means within genotypes. B: Proportion of shoot fresh mass between the treatments (ARD / gARD) for each genotype. C: Comparison of fresh mass proportions between genotypes. Black bars: 95 % confidence intervals (mvt-adjusted for the tests visualised in B and C)."}
# Graphical display of the results

# Observations, means, 95% pointwise CI
plot_SFM_1_treat <- ggplot(dat, aes(x=Treat, y=SFM))+
        theme_bw()+
        geom_jitter(aes(color=Soil),
                    width=0.2,
                    alpha=0.3)+
        geom_pointrange(data=means_SFM_treat,
                        aes(y=response,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        facet_grid( ~ Genotype)+
        # scale_color_manual(values=c("Hg"="#DFA398",
        #                             "Ho"="#883b41",
        #                             "El"="#41917C",
        #                             "Pi"="#A9D284",
        #                             "Ru"="#505778",
        #                             "Me"="#A0A5C0"))+
        ylab("Shoot fresh mass [g]")+
        xlab("Treatment")+
        ggtitle("A: Treatment means")


# Simultaneous 95 % CI for contrast G- ARD within each genotype
ci_SMF_treat <- data.frame(confint(comp_SFM_treat$contrasts))

plot_SFM_2_treat <- ggplot(ci_SMF_treat, aes(x=Genotype))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept=1)+
        xlab("Contrast")+
        ylab("Estimate")+
        coord_flip()+
        ggtitle("B: ARD/G within genotypes")


plot_SFM_3_treat <- ggplot(inter_cont_ci_treat, aes(x=genotype_pw, y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        geom_hline(yintercept=1)+
        ylab("Estimate")+
        xlab("Contrast")+
        coord_flip()+
        ggtitle("C: ARD/G between genotypes")


ggarrange(plot_SFM_1_treat,
          ggarrange(plot_SFM_2_treat,
                    plot_SFM_3_treat,
                    widths = c(2:2.8)),
          nrow=2)
```

\newpage


## Soil treatment interaction

```{r cache=TRUE}
# Treatments within soils
comp_SFM_soil <- emmeans(fit_SFM, 
                    specs="Treat", 
                    by="Soil",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```

```{r}
# Means and their pointwise 95% confidence intervals
means_SFM_soil <- data.frame(comp_SFM_soil$emmeans)
means_SFM_soil
```


```{r}
# Contrasts, p-values etc.
contr_SFM_soil <- data.frame(comp_SFM_soil$contrasts)
contr_SFM_soil
```

\newpage

```{r cache=TRUE}
# Interaction contrasts (ARD / G) between genotypes
# estimates, p-values etc.
inter_means_soil <- emmeans(fit_SFM, specs=c("Treat", "Soil"))

inter_cont_soil <- contrast(inter_means_soil,
                       interaction=c(Treat="pairwise", Soil="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_soil
```

\newpage

```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (ARD / G) between genotypes
inter_cont_ci_soil <- confint(inter_cont_soil) %>% 
        data.frame() %>% 
        mutate(treat_pw=Treat_pairwise,
               soil_pw=Soil_pairwise,
               ratio=round(ratio, 3),
               lower.CL=round(lower.CL, 3),
               upper.CL=round(upper.CL, 3)) %>% 
        select(treat_pw,    
               soil_pw,
               ratio,
               lower.CL,
               upper.CL)

inter_cont_ci_soil
```

### Overview about results

The analysis of the soil-treatment interaction is summarized in Fig. \ref{fig_soil_treat}.
Fig \ref{fig_soil_treat}A provides an overview about the raw data, the average 
shoot fresh mass per treatment within each soil and corresponding pointwise 95 % confidence
intervals.
The proportional change of shoot freshmass between ARD and gamma irradiated substrate
for each soil is given in fig \ref{fig_soil_treat}B together with
multiplicity adjusted 95 % confidence intervals. Since the null hypothesis
($H_0: \mu_i / \mu_{i'} = 1$) is covered by none of the confidence intervals,
it can be deduced that in each soil, the shoot fresh mass for ARD infected 
substrate is significantly lower than in its gamma irradiated counterpart.
The magnitude of shoot fresh mass reduction (ARD/G) is significantly bigger in 
Pinneberg than in Ruthe and the shoot fresh mass in CO is significantly higher
than in RU, He and El. This is because the corresponding confidence intervals are 
relatively far above the 1 (\ref{fig_soil_treat}C). 
The magnitude of shoot fresh mass reduction is significantly smaller in HE than in Pi, 
in He than in Me, in El than in PI, because the complete confidence interval is lower than 
the null hypothesis of 1. Also the magnitude of shoot fresh mass reduction in El
appears to be  smaller than in Me, but the corresponding upper confidence limit 
is slightly above 1.  

```{r echo=FALSE, cache=TRUE, fig.width=10, fig.height=9, fig.cap = "\\label{fig_soil_treat} Analysis of the soil treatment interaction. A: Treatment means within soils. B: Proportional change of shoot fresh mass between the treatments (ARD / gARD) for each soil. C: Comparison of fresh mass proportions between soils. Black bars: 95 % confidence intervals (mvt-adjusted for the tests visualised in B and C)."}
# Graphical display of the results

# Observations, means, 95% pointwise CI
plot_SFM_1_soil <- ggplot(dat, aes(x=Treat, y=SFM))+
        theme_bw()+
        geom_jitter(aes(color=Genotype),
                    width=0.2,
                    alpha=0.3)+
        geom_pointrange(data=means_SFM_soil,
                        aes(y=response,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        facet_grid( ~ Soil)+
        # scale_color_manual(values=c("Hg"="#DFA398",
        #                             "Ho"="#883b41",
        #                             "El"="#41917C",
        #                             "Pi"="#A9D284",
        #                             "Ru"="#505778",
        #                             "Me"="#A0A5C0"))+
        ylab("shoot fresh mass [g]")+
        xlab("Treatment")+
        ggtitle("A: Treatment means")


# Simultaneous 95 % CI for contrast G- ARD within each genotype
ci_SMF_soil <- data.frame(confint(comp_SFM_soil$contrasts))

plot_SFM_2_soil <- ggplot(ci_SMF_soil, aes(x=Soil))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_hline(yintercept=1)+
        xlab("Contrast")+
        ylab("Estimate")+
        coord_flip()+
        ggtitle("B: ARD/G within genotypes")


plot_SFM_3_soil <- ggplot(inter_cont_ci_soil, aes(x=soil_pw, y=ratio))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymax=upper.CL,
                            ymin=lower.CL))+
        geom_hline(yintercept=1)+
        ylab("Estimate")+
        xlab("Contrast")+
        coord_flip()+
        ggtitle("C: ARD/G between genotypes")


ggarrange(plot_SFM_1_soil,
          ggarrange(plot_SFM_2_soil,
                    plot_SFM_3_soil,
                    widths = c(2:2.8)),
          nrow=2)
```

\newpage


## Soil Genotype interaction


```{r cache=TRUE}
# Treatments within soils
comp_SFM_sg <- emmeans(fit_SFM, 
                    specs="Genotype", 
                    by="Soil",
                    contr="pairwise",
                    adjust="mvt",
                    type="response")
```

```{r}
# Means and their pointwise 95% confidence intervals
means_SFM_sg <- data.frame(comp_SFM_sg$emmeans)
means_SFM_sg
```


```{r}
# Contrasts, p-values etc.
contr_SFM_sg <- data.frame(comp_SFM_sg$contrasts)
contr_SFM_sg %>% 
        mutate(ratio=round(ratio, 3),
               SE=round(SE, 3),
               df=round(df, 3),
               t.ratio=round(t.ratio, 3),
               p.value=round(p.value, 3))
```

\newpage

```{r cache=TRUE}
# Interaction contrasts (soil_i / soil_i') between genotypes
# estimates, p-values etc.
inter_means_sg <- emmeans(fit_SFM, specs=c("Genotype", "Soil"))

inter_cont_sg <- contrast(inter_means_sg,
                       interaction=c(Genotype="pairwise", Soil="pairwise"),
                       adj="mvt",
                       type="response") 
inter_cont_sg
```



\newpage

```{r cache=TRUE}
# Adjusted 95% CI for interaction contrasts
# (soil_i / soil_i') between genotypes
# Takes very long computing time
# inter_cont_ci_sg <- confint(inter_cont_sg) %>%
#         data.frame() %>%
#         mutate(treat_pw=Treat_pairwise,
#                soil_pw=Soil_pairwise,
#                ratio=round(ratio, 3),
#                lower.CL=round(lower.CL, 3),
#                upper.CL=round(upper.CL, 3)) %>%
#         select(treat_pw,
#                soil_pw,
#                ratio,
#                lower.CL,
#                upper.CL)
# 
# inter_cont_ci_soil
```

# Threeway interaction (Soil:Genotype:Treat)

```{r cache=TRUE}
# Check if ARD/G differs from 1 in which genotype:soil combination
comp_SFM_3 <- emmeans(fit_SFM, 
                    specs="Treat", 
                    by=c("Soil", "Genotype"),
                    contr="pairwise",
                    adjust="mvt",
                    type="response")

contr_3 <- data.frame(comp_SFM_3$contrasts)

contr_3 %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               p.value=round(p.value, 4)) %>% 
        select(contrast,
               Soil,
               Genotype,
               ratio,
               SE,
               df,
               p.value)
```

```{r fig.width=10, fig.height=10, fig.cap = "\\label{fig_3_way} Estimated proportions for ARD/G for each combination of soil and genotype."}
contr_3_ci <- data.frame(confint(comp_SFM_3$contrasts))

ggplot(contr_3_ci,aes(x=contrast))+
        theme_bw()+
        geom_pointrange(aes(y=ratio,
                            ymin=lower.CL,
                            ymax=upper.CL))+
        geom_point(data=contr_SFM_soil,
                   aes(y=ratio,
                       x=contrast),
                   color="red",
                   alpha=0.6)+
        geom_point(data=contr_SFM_treat,
                   aes(y=ratio,
                       x=contrast),
                   color="blue",
                   alpha=0.6)+
        facet_grid(Genotype~Soil)+
        geom_hline(yintercept=1)+
        xlab("Contrast")+
        ylab("Estimate")
```

\newpage

For all combinations of the soils and genotypes plants grown in ARD infected substrate
yield less shoot fresh mass than plants in gamma irradiated substrate (fig. \ref{fig_3_way}).
However, the growthreduction in soil Pi is not significant for genotype G202. Also
in soil Co the growthreduction is not significantly different for the genotypes
G935 and Mal1030. All other differences between ARD and gamma irradiated substrate
are found to be significant, since the corresponding confidence intervals do not 
cover the 1. None of the interaction contrasts given below is significant.  

<!-- Hence, the reduction of shoot fresh mass caused by the ARD infection (compared to gamma  -->
<!-- irradiated substrate) is constant for all genotype-soil combinations and the -->
<!-- three-way interaction is found to be absent (as already shown in the ANOVA table). -->


```{r cache=TRUE}
# Interaction contrasts
means_3 <- emmeans(fit_SFM,
                   specs= ~ Soil:Genotype:Treat,
                    type="response")

inter_cont_3 <- contrast(means_3,
                       interaction=c(Genotype="pairwise", 
                                     Soil="pairwise",
                                     Treat="pairwise"),
                       adj="mvt",
                       type="response") 

inter_cont_3_df <- data.frame(inter_cont_3)

inter_cont_3_df %>% 
        mutate(ratio=round(ratio, 2),
               SE=round(SE, 2),
               df=round(df, 2),
               p.value=round(p.value, 2)) %>% 
        select(Treat_pairwise,
               Soil_pairwise,
               Genotype_pairwise,
               ratio,
               SE,
               df,
               p.value)

```


